---
title: "R graphing basics"
author: "L.Lofgren & D.Carterhouse"
date: "2/10/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 
  R is a language that excells at plotting and statistical analysis 
  Today we're going to talk about plotting, and get some hands-on practice at plotting and problem selving. 
  
  We'll start by some basic plot types and compare/contrast base and gg-plot
  
  NOTE: We are going to make some assumptions here about getting your data into R in the first place (this should be it's own  workshop). 
  
  Today we'll use a standardized dataset that's preloaded in R's envronment:
  (Orange) - demo only
  (Cabbage) - used in workshop 

---

## If you get lost 


This markdown document is avalible on googledrive 

If you need help with syntax:

help(<plot-type>)

Help-> Cheat Sheets

---

## Plot types

We're going to go over a few basic plot types:

- Strip charts 
- Boxplots
- Scatter plots

We're going to render each plot type in two ways, useing base and gg-plot

---

## What's the difference? 
 - **base:** base R uses only functions that come preinstalled into R's envronement  
    *pros:* 
    + highly flexible    
    
    *cons:* 
    + learning curve is greater     
    + grammer is inconsistant between plot types  
    
 ---

 - **ggplot:** is a graphics package (an extentsion) for R   
    *pros:*  
    + easier to hit the ground running   
    + consistant grammar   
    + ease of multi-paneling  
    + ease of pre-packaged color libraries   
    
    *cons:* 
     + not flexible   
     + kinda dorky
        
---


## get the data
```{r get the data}

#load libraries
library(ggplot2)

#load the data: 
data()
data(Orange)

#get info about your data

#View(Orange)
#?Orange
str(Orange)
head(Orange)
```

---

## Strip charts
- Base
```{r strip charts}
# rememebr if you need help:
help(stripchart)

stripchart(Orange$age)

```

---

```{r}

stripchart(Orange$circumference ~ Orange$Tree,
           method="jitter", 
           vertical=TRUE,
           main='Oranges',
           xlab='Tree ID',
           ylab='Tree circumference',
           pch = 16, 
           col = c("#326773", "#557329", "#A68217", "#D9751E", "#F2622E"))

```

--- 

```{r}

Orange$Tree
Orange$Tree <- factor(Orange$Tree, levels = c(1, 2, 3, 4, 5))
```

---

```{r}

help(sort)
stripchart(Orange$circumference ~ Orange$Tree,
           method="jitter", 
           vertical=TRUE,
           main='Oranges',
           xlab='Tree ID',
           ylab='Tree circumference',
           pch = 16, 
           col = c("#326773", "#557329", "#A68217", "#D9751E", "#F2622E"))
```

---

## the anatomy of a gg-plot  
gg-calls this the "grammar of graphics"

- gg = "grid graphics"

- devided into blocks of arguments 
  *  data: a data frame: categorical, numeric etc.
  *  aesthetic mapping (aec) of variables into visual properties: size, color, x, y
  *  geometric objects (geom): points, lines, areas, arrows, …
  *  coordinate system (coord): Cartesian, log, polar, map, 
  
--- 

## example:

```{r gg plot example}
ggplot(data = Orange, 
       aes(x=Tree, y=circumference, color=Tree)) +
 geom_jitter(width = 0.25, size=3) 

```

---

## use theme_minimal()

```{r example w/ minimual}
my_plot<- ggplot(Orange, aes(x=Tree, y=circumference, color=Tree)) +
 geom_jitter(width = 0.25, size=3) 
my_plot + theme_minimal()
 
```

---

## note - you can set minimal as the default like so:

```{r}
library(ggplot2)
theme_set(
  theme_minimal() +
    theme(legend.position = "right")
  )
```

---

Color resources:

https://color.adobe.com/  
R-color brewer  
lots and lots and lots more  

---

## box plots

#base
```{r}
boxplot(circumference~Tree, data = Orange,
        ylim=c(0,max(Orange$circumference)),
        main="Oranges",
        ylab="circumference",
        axis.lty="solid", 
        col = c("#326773", "#557329", "#A68217", "#D9751E", "#F2622E"),
        axisnames = TRUE, 
        cex.names = .4, 
        las =1)

```

---

## show your data
```{r}
stripchart(circumference~Tree, data = Orange, xlim=c(0.5,6.5), vertical=TRUE, method="stack", offset=0.8, pch=19,
main = "Oranges", xlab = "Tree ID", ylab = "circumference")

boxplot(circumference~Tree, data = Orange,
        add=TRUE,
        range=0,
        ylim=c(0,max(Orange$circumference)),
        main="Oranges",
        ylab="circumference",
        axis.lty="solid", 
        col = c("#32677399", "#55732999", "#A6821799", "#D9751E99", "#F2622E99"),
        axisnames = TRUE, 
        cex.names = .4, 
        las =1, 
        #whisklty = 0, 
        staplelty = 0)

```

--- 

## gg-plot
```{r}
my_plot<- ggplot(Orange, aes(x=Tree, y=circumference, color=Tree)) +
  geom_jitter(width = 0.25, size=3, show.legend = FALSE)
my_plot + theme_minimal() + geom_boxplot(aes(alpha = .2), show.legend = FALSE) 
```

---

```{r}
#add grouping variable for strip chart
# "#D9751E","#F2622E")[as.numeric(Orange$Tree)])
```

